\documentclass[12pt]{article}
\usepackage{fullpage,amsmath,amsfonts,mathpazo,microtype,nicefrac}

\usepackage[font=footnotesize,skip=-10pt]{caption}

% Set-up for hypertext references
\usepackage{hyperref,color,textcomp}
\definecolor{webgreen}{rgb}{0,.35,0}
\definecolor{webbrown}{rgb}{.6,0,0}
\definecolor{RoyalBlue}{rgb}{0,0,0.9}
\hypersetup{
   colorlinks=true, linktocpage=true, pdfstartpage=3, pdfstartview=FitV,
   breaklinks=true, pdfpagemode=UseNone, pageanchor=true, pdfpagemode=UseOutlines,
   plainpages=false, bookmarksnumbered, bookmarksopen=true, bookmarksopenlevel=1,
   hypertexnames=true, pdfhighlight=/O,
   urlcolor=webbrown, linkcolor=RoyalBlue, citecolor=webgreen,
   pdfauthor={Chris H. Rycroft},
   pdfsubject={UW--Madison Math 521 (Spring 2025)},
   pdfkeywords={},
   pdfcreator={pdfLaTeX},
   pdfproducer={LaTeX with hyperref}
}
\hypersetup{pdftitle={Math 521: Assignment 4}}

% Macro definitions
\newcommand{\N}{\mathbb{N}}
\newcommand{\B}{\mathbb{B}}
\newcommand{\I}{\mathbb{I}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\p}{\partial}
\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\vu}{\vec{u}}
\newcommand{\vv}{\vec{v}}
\newcommand{\vx}{\vec{x}}
\newcommand{\vy}{\vec{y}}
\newcommand{\vp}{\vec{p}}
\newcommand{\sep}{\,:\,}
\newcommand{\Trans}{\mathsf{T}}

\begin{document}
\section*{Math 521: Assignment 4 (due 5~PM, March 21)}
\begin{enumerate}
  \item Consider the metric space $(\Q,d)$ with $d(x,y) = |x-y|$ and define
    $F=\{ p\in \Q \sep 2<p^2<5\}$. Show that $F$ is closed and bounded in $\Q$
    but that $F$ is not compact. Is $F$ open in $\Q$?
  \item If $(S,d)$ is a metric space, prove that the finite union of compact
    sets in $S$ is compact. Construct an example to show this does not
    generalize to infinite unions.
  \item Let $(S,d_S)$ and $(T,d_T)$ be two metric spaces. Define a product set
    $S\times T=\{(s,t) \sep s\in S, t\in T\}$. For two elements $(s_1,t_1),
    (s_2,t_2)\in S\times T$, define the metric
    \begin{equation}
      d_{S\times T}((s_1,t_1),(s_2,t_2)) = d_S(s_1,s_2) + d_T(t_1,t_2).
    \end{equation}
    Let $K \subseteq S$ and $L \subseteq T$, and define $K\times L = \{(s,t)
    \sep s\in K, t\in L\}$. Show that if $K\times L$ is compact, then $K$ and
    $L$ are also compact. \textbf{Optional:} is the converse result true?
  \item Determine the interior and closure of the following subsets of $\R$:
    \begin{equation}
      A= (-1,1] \cap \Q, \qquad B = (-1,1] \cup \Q.
    \end{equation}
  \item Consider the function
    \begin{equation}
      h(x) = \begin{cases}
        1-x^2 & \qquad \text{for $x\in \Q$,} \\
        0 & \qquad \text{for $x\notin \Q$.}
      \end{cases}
    \end{equation}
    Show that $h$ is continuous at $\pm 1$ but at no other points.
  \item For $\alpha \in \R$, define
    \begin{equation}
      f(x) = \begin{cases}
        |x|^\alpha \sin \tfrac1x & \qquad \text{for $x\ne 0$,} \\
        0 & \qquad \text{for $x=0$.}
      \end{cases}
    \end{equation}
    Find the exact range of values of $\alpha$ for which $f$ is continuous
    at 0.
  \item
    \begin{enumerate}
      \item Let $f$ be a continuous real-valued function on an interval
        $(a,b)$. Show that if $f(q)=0$ for all rational numbers $q$ in $(a,b)$,
        then $f(x)=0$ for all $x\in (a,b)$.
      \item Let $f$ and $g$ be continuous real-valued functions on an interval
        $(a,b)$. Show that if $f(q)=g(q)$ for all rational numbers $q$ in
        $(a,b)$, then $f(x)=g(x)$ for all $x\in (a,b)$.
      \item Do the results from parts (a) and (b) still hold for a closed
        interval $[a,b]$?
    \end{enumerate}
  \item Let $f$ be a real-valued function on $\R$. Suppose that for a given
    $x \in \R$,
    \begin{equation}
      \lim_{n\to\infty} \left( f(x+a_n) - f(x-a_n) \right) =0
    \end{equation}
    for all sequences $(a_n)$ that converge to 0. Is $f$ continuous at $x$?
  \item Suppose that $f$ is a real-valued continuous function on $\R$ and
    $f(a)f(b)<0$ for some $a,b\in \R$. Prove that there exists an $x$ between
    $a$ and $b$ such that $f(x)=0$.
  \item Let $f$ be a continuous function on $[a,b]$. Show that the function
    $f^*$ defined as $f^*(x) = \sup \{f(z) \sep a \le z \le x\}$ is an
    increasing continuous function on $[a,b]$.
  \item \textbf{Optional.} A real-valued function on an interval $I$ is called
    convex if for all $x,y\in I$ and $\lambda \in (0,1)$, then
    \begin{equation}
      f((1-\lambda)x+\lambda y) \ge (1-\lambda) f(x) + \lambda f(y).
    \end{equation}
    Suppose that $f$ is convex on $[a,b]$. Prove that $f$ is continuous on
    $(a,b)$, but need not be continuous at $a$ or $b$.    
\end{enumerate}
\end{document}
