\documentclass[12pt]{article}
\usepackage{fullpage,amsmath,amsfonts,mathpazo,microtype,nicefrac}

\usepackage[font=footnotesize,skip=-10pt]{caption}

% Set-up for hypertext references
\usepackage{hyperref,color,textcomp}
\definecolor{webgreen}{rgb}{0,.35,0}
\definecolor{webbrown}{rgb}{.6,0,0}
\definecolor{RoyalBlue}{rgb}{0,0,0.9}
\hypersetup{
   colorlinks=true, linktocpage=true, pdfstartpage=3, pdfstartview=FitV,
   breaklinks=true, pdfpagemode=UseNone, pageanchor=true, pdfpagemode=UseOutlines,
   plainpages=false, bookmarksnumbered, bookmarksopen=true, bookmarksopenlevel=1,
   hypertexnames=true, pdfhighlight=/O,
   urlcolor=webbrown, linkcolor=RoyalBlue, citecolor=webgreen,
   pdfauthor={Chris H. Rycroft},
   pdfsubject={UW--Madison Math 521 (Spring 2025)},
   pdfkeywords={},
   pdfcreator={pdfLaTeX},
   pdfproducer={LaTeX with hyperref}
}
\hypersetup{pdftitle={Math 521: Assignment 3}}

% Macro definitions
\newcommand{\N}{\mathbb{N}}
\newcommand{\B}{\mathbb{B}}
\newcommand{\I}{\mathbb{I}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\p}{\partial}
\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\vu}{\vec{u}}
\newcommand{\vv}{\vec{v}}
\newcommand{\vx}{\vec{x}}
\newcommand{\vy}{\vec{y}}
\newcommand{\vp}{\vec{p}}
\newcommand{\sep}{\,:\,}
\newcommand{\Trans}{\mathsf{T}}

\begin{document}
\section*{Math 521: Assignment 3 (due 5~PM, March 7)}
\begin{enumerate}
  \item Suppose that $0 \le a_n < 1$ for all $n\in \N$. Prove that if
    $\sum_{n=1}^\infty a_n$ converges, then $\sum_{n=1}^\infty a_n^2$ and
    $\sum_{n=1}^\infty a_n/(1-a_n)$ also converge. Are the converse statements
    true?
  \item Let $(u_n)$ and $(v_n)$ be sequences of positive real numbers for $n\in
    \N$. For each of the following statements, either prove it or provide a
    counterexample.
    \begin{enumerate}
        \item If $(u_n)$ and $(v_n)$ are equal except at finitely many $n$, then
      $\sum u_n$ and $\sum v_n$ either both converge or both diverge.
        \item If $(u_n)$ and $(v_n)$ are equal at infinitely many $n$, then $\sum
      u_n$ and $\sum v_n$ either both converge or both diverge.
        \item If $\sum u_n$ and $\sum v_n$ diverge, then $\sum u_n v_n$ diverges.
        \item If $(u_n/v_n) \to 1$ as $n \to \infty$, then $\sum u_n$ and $\sum
      v_n$ both converge or both diverge.
        \item If $u_n-v_n\to 0$, then $\sum u_n$ and $\sum v_n$ both converge
      or both diverge.
        \item If $(u_{n+1}/u_n)>k>1$ for infinitely many $n$, then $\sum u_n$
      diverges.
    \end{enumerate}
  \item Let $s_n$ be the sum of the first $n$ terms of the series
    \begin{equation}
      1 - \frac{1}{\sqrt{2}} + \frac{1}{\sqrt{3}} - \frac{1}{\sqrt{4}} + \frac{1}{\sqrt{5}} - \ldots
    \end{equation}
    and let $t_n$ be the sum of the first $n$ terms of the rearranged series
    \begin{equation}
      1 + \frac{1}{\sqrt{3}} - \frac{1}{\sqrt{2}} + \frac{1}{\sqrt{5}} + \frac{1}{\sqrt{7}} - \frac{1}{\sqrt{4}} + \ldots,
    \end{equation}
    which is formed by taking two positive terms and then one negative term.
    Prove that $(s_n)$ converges. Prove that $t_{3n} \ge s_{2n} + n/\sqrt{4n-1}$
    and hence that $(t_n)$ diverges.
  \item
    \begin{enumerate}
      \item Let $(a_n)$ be a decreasing sequence of positive numbers. Prove
        that the series
        \begin{equation}
          \sum_{n=1}^\infty a_n  \qquad \text{and} \qquad \sum_{n=1}^\infty 2^n a_{2^n}
        \end{equation}
        both converge or both diverge.
      \item Using part (a), prove that $\sum_{n=1}^\infty n^{-\alpha}$ is convergent if and only if $\alpha>1$.
      \item Prove that $\sum_{n=2}^\infty n^{-1} (\log n)^{-\alpha}$ is convergent if and only if $\alpha>1$.
    \end{enumerate}
  \item
    Using the integral test, or otherwise, determine whether the following series converge:
    \[
      \textrm{(a)}\;\sum_n \frac{1}{\sqrt{n}\log n}, \quad
      \textrm{(b)}\;\sum_n \frac{1}{n(\log n)(\log(\log n))}, \quad
      \textrm{(c)}\;\sum_n \frac{\log n}{n}, \quad
      \textrm{(d)}\;\sum_n \frac{\log n}{n^2}.
    \]
  \item \textbf{Optional.} Find a sequence $(a_n)$ such that
    \begin{equation}
      \lim_{n\to\infty} a_n n(\log n)(\log(\log n)) =0
    \end{equation}
    but $\sum a_n$ diverges.
  \item Consider the functions defined for $x,y\in \R$:
    \[
      d_1(x,y) = (x-y)^2, \qquad d_2(x,y) = \sqrt{|x-y|}, \qquad d_3(x,y) = |x^3-y^3|,
    \]
    \[
      d_4(x,y) = |x^4 - y^4|, \qquad d_5(x,y) = \min \{ |x-y|, 1\}.
    \]
    For each function, determine whether it is a metric or not. For the
    functions that are metrics, calculate the neighborhood $N_2(1)$.
  \item Define the distance function
    \begin{equation}
      d(x,y) = \begin{cases}
        |x|+ |y| & \qquad \text{for $x\ne y$,} \\
        0 & \qquad \text{for $x=y$.}
      \end{cases}
    \end{equation}
    Prove that $d$ is a metric on $\R$. For $0<a<b$, show that the
    open interval $(a,b) = \{x \in \R \sep a<x<b\}$ is both open
    and closed with respect to $d$. Determine whether or not $(\R,d)$
    is a complete metric space.
  \item Let $B$ be the set of all bounded sequences $\vu=(u_1,u_2,\ldots)$. For
    each of the following functions, prove that it is a metric, or explain why
    it is not.
    \begin{enumerate}
      \item $d(\vu,\vv)=\sup\{|u_j - v_j| \sep j \in \N\}$.
      \item $d(\vu,\vv)=\sum_{j=1}^\infty 2^{-j} |u_j - v_j|$.
      \item $d(\vu,\vv)=\sum_{j=1}^\infty \tfrac{1}{j} |u_j -v_j|$.
    \end{enumerate}
    \textbf{Optional.} For the three different definitions of $d$, determine
    whether or not $(B,d)$ is a complete metric space.
  \item Consider $\R^2$ with positions written as $\vx=(x_1,x_2)$, and
    define the Euclidean norm as $\|\vx\| = (x_1^2+x_2^2)^{1/2}$.
    The Poincar\'e disk consists of the points $S=\{\vx \sep \|\vx\|<1\}$
    with metric
    \begin{equation}
      d(\vx,\vy) = \cosh^{-1} \left( 1+ \frac{2\|\vx-\vy\|^2}{(1-\|\vx\|^2) (1-\|\vy\|^2)} \right)
    \end{equation}
    for all $\vx,\vy\in S$. Define $r=\cosh^{-1} \tfrac{3}{2}$. Draw the
    Poincar\'e disk, and then calculate and draw the neighborhoods $N_r(\vx)$
    for $\vx=(0,0)$, $(\tfrac12,0)$, and $(\tfrac34,0)$.\footnote{This can be
    done analytically, although you may use a computer if you wish.}
  \item Let $(p_n)$ be a Cauchy sequence in a metric space $(S,d)$.
    Suppose that a subsequence $(p_{n_k})$ converges to a point $p\in S$.
    Prove that the full sequence converges to $p$.
\end{enumerate}
\end{document}
